A matrix-free approach for finite-strain hyperelastic problems using geometric multigrid

### 2018-10-18 doc/large_strain.tex
more

%
% Deformation gradient
%
The tangent map is linear such that
$\d \gz x = \gz F \cdot \d \gz X$,
where $\gz F := \Grad \, \gz \varphi \equiv \gz I + \Grad \, \gz u$ is called deformation gradient and $\gz I$ is the second order unit tensor.
The mapping has to be one-to-one and must exclude self-penetration. Consequently, the Jacobian $J = \det \gz F > 0$ has to be positive.
% The deformation gradient can be split into the proper orthogonal rotation tensor $\gz R$ and the symmetric right $\gz U$ and left $\gz V$ stretch tensors such that $\gz F = \gz R \gz U = \gz V \gz R$.
We shall denote the push-forward transformation of rank-2 and rank-4 tensors $\mathbf{A}$ and $\boldsymbol{\mathcal{A}}$ by
\begin{align}
  \varphi\left( \mathbf{A} \right)_{ij}
  &= F_{iA} A_{AB} F_{jB} \\
\varphi\left( \boldsymbol{\mathcal{A}} \right)_{ijkl}
  &= F_{iA} F_{jB} \mathcal{A}_{ABCD} F_{kC} F_{lD}.
\end{align}

@@

In order to solve \eqref{eq:weak_form} using Newton's method, a first order approximation around a given solution field $\overline{\gz u}$ is required such that
\begin{align}
F(\overline{\gz u} + \Delta \gz u, \delta \gz u) \approx
F(\overline{\gz u}, \delta \gz u) + \D_{\Delta \gz u} F(\overline{\gz u}, \delta \gz u).
\end{align}
For conservative traction boundary conditions, the directional derivative is given by
\begin{align}
\D_{\Delta \gz u}  F(\overline{\gz u}, \delta \gz u)
=
\int_{\mcl B_0} \D_{\Delta \gz u} \gz \tau : \overline{\grad^s} \, \delta \gz u \, \d V
+
\int_{\mcl B_0} \overline{\gz \tau} : \left[ \Grad \, \delta \gz u \cdot \D_{\Delta \gz u}  \gz F^{-1} \right] \d V.
\end{align}
Following \cite{Wriggers2008}, this can be simplified to
\begin{align}
\D_{\Delta \gz u} F(\overline{\gz u}, \delta \gz u)
  =
  \int_{\mcl B_0} \overline{\grad^s} \Delta \gz u : J \boldsymbol{\mathcal{C}} : \overline{\grad^s} \, \delta \gz u \, \d V
  +
  \int_{\mcl B_0}
  \overline{\grad}\delta \gz u :
  \left[
  \overline{\grad} \Delta \gz u \cdot
  \overline{\gz \tau}
  \right]
  \d V.
\label{eq:tangent}
\end{align}
Here $\overline{(\bullet)}$ is used to denote quantities evaluated using the displacement field $\overline {\gz u}$, and the fourth order tangent tensor is the push forward of the material tangent $J \boldsymbol{\mathcal{C}} = \varphi\left( 4 \frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}} \right)$. The first term in \eqref{eq:tangent} is related to the linearization of the Kirchhoff stress and therefore called material part of the tangent operator.  The second term in \eqref{eq:tangent} is called geometric part of the tangent since it originates from the linearization of the inverse of the deformation gradient $\gz F^{-1}$.

@@

Hence, we can expect that such a matrix-free approach is faster than a regular matrix-based approach, especially for 3D and high polynomial degree with the consistent quadrature rule.
More information on techniques developed to increase the computational density can be found in \cite{kronbichler12,vos10}.

@@

Given that the matrix-free operators are applied on-the-fly, it is crucial to find a good balance between pre-calculation of certain values (caching) and performance of the algorithm.
Below, we propose three algorithms to implement tangent operator of finite-strain linear elasticity (see \eqref{eq:algebraic_tangent}) using the matrix-free approach.
%
In order to obtain gradients with respect to the deformed configuration we used the \textit{MappingQEulerian} class from the \texttt{deal.II} \cite{dealII90} library.
In addition to the standard mapping from the reference (isoparametric) element to the element in real space, this class computes the mapping to the deformed configuration given a displacement field.
%
Due to the specifics of matrix-free operator evaluation implemented in \texttt{deal.II}, the integration is performed over the deformed configuration. Therefore, we have to divide by the Jacobian $J$ of the mapping at a given quadrature point additionally to express the integral in the undeformed configuration.

@@

Algorithm \ref{alg:mf_scalar} caches a single scalar which involves the logarithm of the Jacobian -- a computationally demanding operation compared to additions and multiplications. As evident from Section \ref{sec:theory}, we need to reevaluate the Kirchhoff stress at each quadrature point in order to apply the operator. This in turn requires evaluating the deformation gradient $\gz F$ and therefore gradients with respect to the undeformed configuration.

\begin{algorithm}[!ht]
  \SetKwInOut{Input}{Given}
  \SetKwInOut{Output}{Return}
  \Input{RHS FE vector $\gz x$,
  cached
  $c_1 := 2\left[\mu - 2 \lambda \log(J)\right]/J$,
  $c_2 := 2\lambda/J$ and $\gz \tau/J$ for each cell and quadrature point,
  evaluated based on $\overline{\gz u^h}$}
  \Output{action of the FE tangent \eqref{eq:algebraic_tangent} on $\gz x$}
  \ForEach{ element $K \in \Omega^h$ }{
        \ForEach{quadrature point $q$ on $K$}{
          evaluate $\gz g := \gz \nabla \gz x$ \tcp*{2nd order}
          evaluate $\boldsymbol{\mathcal{G}}\gz g = \gz g \cdot \left[\gz \tau/J \right]$ \tcp*{2nd order}
          queue $\boldsymbol{\mathcal{G}} \gz g$ for contraction $\gz \nabla \gz N_i : \boldsymbol{\mathcal{G}} \gz g$ \;
          evaluate $\gz g_s := \gz \nabla_{sym} \gz x$ \tcp*{2nd order symmetric}
          evaluate $\boldsymbol{\mathcal{C}} \gz g_s = c_1 \gz g_s + c_2 \, \rm{tr}(\gz g_s) \gz I$ \tcp*{2nd order symmetric}
          queue $\boldsymbol{\mathcal{C}} \gz g_s$ contraction $\gz \nabla_{sym} \gz N_i : \boldsymbol{\mathcal{C}} \gz g_s $ \;
        }
        evaluate queued contractions using sum factorization and distribute results to the destination vector
  }
  \caption{Matrix-free tangent cell operator, variant 1: cache second order Kirchhoff stress $\gz \tau$ and thereby avoid the need to evaluate refential quantities like $\gz F$ at runtime.}
  \label{alg:mf_tensor2}
\end{algorithm}

In order to avoid recalculation of the Kirchhoff stress, Algorithm \ref{alg:mf_tensor2} caches its value for each element and quadrature point. In this case, we also avoid the need to evaluate gradients of displacement field with respect to the undeformed configuration.
This algorithm also utilizes the chosen constitutive relationship in that the operation of $J \boldsymbol{\mathcal{C}}$ remains directly expressed in the operator itself. It is, however, somewhat more generic than Algorithm \ref{alg:mf_tensor1} since the dependence of $\gz \tau$ exists outside of the oeperator itself (although it must remain consistent with the action of $J \boldsymbol {\mathcal{C}}$).
Therefore, the action of the material fourth order tangent $J \boldsymbol{\mathcal{C}}$ on the second order symmetric tensor $\gz g_s$, as defined in \eqref{eq:simplified_action}, is cheap to evaluate.
To that end, we cache two scalars that depend on the Jacobian $J$.

\begin{algorithm}[!ht]
  \SetKwInOut{Input}{Given}
  \SetKwInOut{Output}{Return}
  \Input{RHS FE vector $\gz x$,
  cached $\gz \tau/J$ and $\boldsymbol{\mathcal{C}}$ for each cell and quadrature point,
  evaluated based on $\overline{\gz u^h}$}
  \Output{action of the FE tangent \eqref{eq:algebraic_tangent} on $\gz x$}
  \ForEach{ element $K \in \Omega^h$ }{
        \ForEach{quadrature point $q$ on $K$}{
          evaluate $\gz g := \gz \nabla \gz x$ \tcp*{2nd order}
          evaluate $\boldsymbol{\mathcal{G}} \gz g = \gz g \cdot \left[\gz \tau/J\right]$ \tcp*{2nd order}
          queue $\boldsymbol{\mathcal{G}} \gz g$ for contraction $\gz \nabla \gz N_i : \boldsymbol{\mathcal{G}} \gz g$\;
          evaluate $\gz g_s := \gz \nabla_{sym} \gz x$ \tcp*{2nd order symmetric}
          evaluate $\boldsymbol{\mathcal{C}}\gz g_s = \boldsymbol{\mathcal{C}}\gz : \gz g_s$ \tcp*{2nd order symmetric}
          queue $\boldsymbol{\mathcal{C}}\gz g_s$ for contraction $\gz \nabla_{sym} \gz N_i : \boldsymbol{\mathcal{C}}\gz g_s$ \;
        }
        evaluate queued contractions using sum factorization and distribute results to the destination vector
  }
  \caption{Matrix-free tangent cell operator: cache fourth order material tangent $\boldsymbol{\mathcal{C}}$ and Kirchhoff stress $\gz \tau$.}
  \label{alg:mf_tensor4}
\end{algorithm}

Finally, Algorithm \ref{alg:mf_tensor4} represents the most general case which does not assume any form of the material tangent tensor.
In this case, we have to cache the fourth order symmetric tensor $\boldsymbol{\mathcal{C}}$ and Kirchhoff stress $\gz \tau$ for each element and quadrature point, and perform the double contraction with the second-order symmetric tensor $\gz g_s$ on-the-fly.

@@

For this work we adopt the matrix-free level operators within the GMG and employ Chebyshev pre- and post-smoothing on each level. In general, given a heterogeneous non-linear material the definition of transfer operators (for restriction and prolongation) and level operators is not trivial. One approach is to start from an arbitrary heterogeneous material at the fine scale and employ homogenization theory \cite{Suquet1985, Hill1972,Hashin1983,Castaneda1997} to design suitable transfer operators \cite{Miehe2007}\footnote{
  We note that this approach is limited to the case where for each quadrature point within an element the fourth-order tangent operator is the same.
}.
In this approach, the fine-scale displacement is decomposed into long-wave and short-wave contributions by an additive split, where the former is associated to the homogeneous response of a patch of elements and the latter represents fluctuations due to the heterogeneous structure of the patch. However, for a patch of elements consisting of the same material, the transfer operators reduce to the standard geometric transfer. Based on this observation we adopt the following approach: It is assumed that the mesh at the coarsest level can accurately describe the heterogeneous fine-strain elastic material. In this case, the restriction and prolongation operations for patches of elements are always performed for the same material and thus admits usage of standard geometric transfer operators. We define the level operators as follows: Recall that the fine scale tangent \eqref{eq:algebraic_tangent} is obtained by linearization around the current displacement field $\gz u^h$. We then restrict this displacement field to all multigrid levels $\{l\}$ and evaluate tangent operators using the matrix-free algorithms from Section \ref{sec:mf}. Essentially, level operators correspond to the linearization around smoothed representation of displacement field\footnote{
Note that this differs from the classical approach where the (tangent) operator $A^{l+1}$ on level $l+1$ is directly related to the (tangent) operator $A^{l}$ on level $l$ via $A^{l+1}=I^{l+1}_{l}  A^l I^l_{l+1}$, where $I^l_{l+1}$ and $I^{l+1}_l$ are global prolongation (coarse-to-fine) and restriction (fine-to-coarse) operators, respectively.
}
Finally, at the coarsest level we use a CG solver to reduce the residual by three orders of magnitude within (at most) one hundred steps.
The multigrid algorithm is a V-cycle with two pre- and post-smoothing steps. For other approaches that combine homogenization and multigrid methods for small strain elastic materials, see \cite{Bayreuther2003,Fish1995}.

@@

In this section, we apply the proposed matrix-free operator algorithms for finite-strain hyperelastic material as well as the geometric multigrid preconditioner to a benchmark problem from \cite{Miehe2007}. The coarse mesh for the 2D and 3D problems is illustrated in Figure \ref{fig:miehe}. The 2D material consists of a square (depicted in blue), a hole and two spherical inclusions (depicted in red). The size of the square domain is $10^{-3} \rm{mm}$.
The matrix material is taken to have Poisson's ratio $0.3$ and shear modulus $\mu=0.4225 \times 10^6\, \rm{N/mm^2}$. The inclusion is taken to be $100$ times stiffer.
The 3D material is obtained by extrusion of the 2D geometry into the third dimension.
Note that, although a relatively coarse mesh is used at the coarsest multigrid level, the geometry of the inclusions is captured accurately by manifold descriptions of the boundaries and interfaces.
Both domains are fully fixed along the bottom surface and a distributed load is applied at the top in the $(1,0)$ or $(1,1,0)$ direction for the 2D and the 3D problems, respectively.
The distributed load of $0.5\,\rm{N}$ is applied in 5 steps.
With respect to the nonlinear solver, the displacement tolerance of the $\mathcal{\ell}_2$ norm for the Newton update is taken to be $10^{-5}$ whereas the relative and absolute tolerances for the forces is $10^{-8}$. The relative convergence criteria for linear solver is $10^{-6}$.

\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{0.8\textwidth}
    \centering
    \includegraphics[width=\textwidth]{gmg_2d.png}
    % \caption{2D multigrid mesh with 2 global refinements}
  \end{subfigure}
  \caption{2D multigrid mesh after two global refinements distributed into three MPI processes. Color indicate the process owning the element.}%
  \label{fig:miehe_gmg}
\end{figure}
%
Figure \ref{fig:miehe_gmg} illustrates the hierarchy of multigrid meshes for the 2D problem after two global refinements.

\begin{table}
  \centering
  \begin{tabular}{ccccc}
  \hline
    $p$ & $q$ & $N_{gref}$ & $N_{el}$ & $N_{DoF}$ \\
  \hline
    1 & 2 & 7 & 1441792 & 2887680 \\
    2 & 3 & 6 & 360448 & 2887680 \\
    3 & 4 & 5 & 90112 & 1625088 \\
    4 & 5 & 5 & 90112 & 2887680 \\
    5 & 6 & 5 & 90112 & 4510720 \\
    6 & 7 & 4 & 22528 & 1625088 \\
    7 & 8 & 4 & 22528 & 2211328 \\
    8 & 9 & 4 & 22528 & 2887680 \\
  \hline
  \end{tabular}
  \caption{Parameters for the 2D benchmark.
  }
  \label{tab:input_parameters_2d}
\end{table}

\begin{table}
  \centering
  \begin{tabular}{ccccc}
  \hline
    $p$ & $q$ & $N_{gref}$ & $N_{el}$ & $N_{DoF}$ \\
  \hline
    1 & 2 & 4 & 1441792 & 4442880 \\
    2 & 3 & 3 & 180224 & 4442880 \\
    3 & 4 & 2 & 22528 & 1891008 \\
    4 & 5 & 2 & 22528 & 4442880 \\
  \hline
  \end{tabular}
  \caption{Parameters for the 3D benchmark.
  }
  \label{tab:input_parameters_3d}
\end{table}

@@

We are interested in the following metrics:
\begin{itemize}
\item the wall-clock time and memory requirement per DoF for matrix-vector multiplication;
\item the average number of CG iterations throughout the entire simulation (i.e. each Newton-Raphson iteration and each loading step);
\item the wall-clock time per DoF for the CG solver;
\end{itemize}
%
The results of the matrix-free approach with a geometric-multigrid preconditioner are compared to the matrix-based approach in conjunction with the algebraic multigrid (AMG) preconditioner using the package ML of the Trilinos \cite{Heroux2005} library, version 12.12.1.
The aggregation threshold for the AMG preconditioner is taken as $10^{-4}$.
Computations are done on a single node of two clusters:
A node on ``Emmy'' cluster has two Xeon 2660v2 ``Ivy Bridge'' chips (10 cores per chip + SMT) running at 2.2 GHz with 25 MB shared cache per chip and 64 GB of RAM. Intel's compiler version 18.03 with flags ``-O3 -march=native'' and Intel MPI version 18.03 are used.
The ``IWR'' results were obtained on a single machine with eight Xeon E7-8870 ``Sandy Bridge'' chips (10 cores per chip  + SMT) running at 2.4 GHz with 30 MB shared cache per chip. For the simulation, GCC version 8.1.0 with flags ``-O3 -march=native'' and OpenMPI version 3.0.0 were used.

@@

\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{0.49\textwidth}
      \centering
      \includegraphics[width=\textwidth]{Emmy_RRZE_timing2d.eps}
      \caption{vmult (2D)}
      \label{fig:benchmark_miehe_Emmy_vmult2}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{Emmy_RRZE_timing3d.eps}
    \caption{vmult (3D)}
    \label{fig:benchmark_miehe_Emmy_vmult3}
  \end{subfigure}
  ~
  \begin{subfigure}[b]{0.49\textwidth}
      \centering
      \includegraphics[width=\textwidth]{Emmy_RRZE_memory2d.eps}
      \caption{memory consumption (2D)}
      \label{fig:benchmark_miehe_Emmy_memory2}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{Emmy_RRZE_memory3d.eps}
    \caption{memory consumption (3D)}
    \label{fig:benchmark_miehe_Emmy_memory3}
  \end{subfigure}
  \caption{Emmy cluster, RRZE, matrix-vector multiplication.}%
  \label{fig:benchmark_miehe_Emmy}
\end{figure}

>>>>

%
% Deformation gradient
%
The tangent map is linear such that
$\d \gz x = \gz F \cdot \d \gz X$,
where $\gz F := \Grad \, \gz \varphi \equiv \gz I + \Grad \, \gz u$ is called deformation gradient and $\gz I$ is the second-order unit tensor.
The mapping has to be one-to-one and must exclude self-penetration. Consequently, the Jacobian $J = \det \gz F > 0$ has to be positive.
% The deformation gradient can be split into the proper orthogonal rotation tensor $\gz R$ and the symmetric right $\gz U$ and left $\gz V$ stretch tensors such that $\gz F = \gz R \gz U = \gz V \gz R$.
We shall denote the push-forward transformation of rank-2 and rank-4 tensors $\mathbf{A}$ and $\boldsymbol{\mathcal{A}}$ by
\begin{align}
  \varphi\left( \mathbf{A} \right)_{ij}
  &= F_{iA} A_{AB} F_{jB} \\
\varphi\left( \boldsymbol{\mathcal{A}} \right)_{ijkl}
  &= F_{iA} F_{jB} \mathcal{A}_{ABCD} F_{kC} F_{lD}.
\end{align}

@@

In order to solve \eqref{eq:weak_form} using Newton's method, a first order approximation around a given solution field $\overline{\gz u}$ is required such that
\begin{align}
F(\overline{\gz u} + \Delta \gz u, \delta \gz u) \approx
F(\overline{\gz u}, \delta \gz u) + \D_{\Delta \gz u} F(\overline{\gz u}, \delta \gz u).
\end{align}
For conservative traction boundary conditions, the directional derivative is given by
\begin{align}
\D_{\Delta \gz u}  F(\overline{\gz u}, \delta \gz u)
=
\int_{\mcl B_0} \D_{\Delta \gz u} \gz \tau : \overline{\grad^s} \, \delta \gz u \, \d V
+
\int_{\mcl B_0} \overline{\gz \tau} : \left[ \Grad \, \delta \gz u \cdot \D_{\Delta \gz u}  \gz F^{-1} \right] \d V.
\end{align}
Following \cite{Wriggers2008}, this can be simplified to
\begin{align}
\D_{\Delta \gz u} F(\overline{\gz u}, \delta \gz u)
  =
  \int_{\mcl B_0} \overline{\grad^s} \Delta \gz u : J \boldsymbol{\mathcal{C}} : \overline{\grad^s} \, \delta \gz u \, \d V
  +
  \int_{\mcl B_0}
  \overline{\grad}\delta \gz u :
  \left[
  \overline{\grad} \Delta \gz u \cdot
  \overline{\gz \tau}
  \right]
  \d V.
\label{eq:tangent}
\end{align}
Here $\overline{(\bullet)}$ is used to denote quantities evaluated using the displacement field $\overline {\gz u}$, and the fourth-order tangent tensor is the push forward of the material tangent $J \boldsymbol{\mathcal{C}} = \varphi\left( 4 \frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}} \right)$. The first term in \eqref{eq:tangent} is related to the linearization of the Kirchhoff stress and therefore called material part of the tangent operator.  The second term in \eqref{eq:tangent} is called geometric part of the tangent since it originates from the linearization of the inverse of the deformation gradient $\gz F^{-1}$.

@@

Hence, we can expect that such a matrix-free approach is faster than a regular matrix-based approach, especially for 3D and high polynomial degree with the corresponding quadrature rule.
More information on matrix-free techniques can be found in \cite{kronbichler12,vos10}.

@@

Given that the matrix-free operators are applied on-the-fly, it is crucial to find a good balance between pre-calculation of certain values (caching) and performance of the algorithm.
Below, we propose three algorithms to implement the tangent matrix of finite-strain elasticity (see \eqref{eq:algebraic_tangent}) using the matrix-free approach.
%
In order to obtain gradients with respect to the deformed configuration we used the \textit{MappingQEulerian} class from the \texttt{deal.II} \cite{dealII90} library.
In addition to the standard mapping from the reference (isoparametric) element to the element in real space, this class computes the mapping to the deformed configuration given a displacement field.
%
Due to the specifics of matrix-free operator evaluation implemented in \texttt{deal.II}, the integration is performed over the deformed configuration. Therefore, we have to divide by the Jacobian $J$ of the deformation map at a given quadrature point additionally to express the integral in the undeformed configuration.

@@

Algorithm \ref{alg:mf_scalar} caches a single scalar which involves the logarithm of the Jacobian -- a computationally demanding operation compared to additions and multiplications. As evident from Section \ref{sec:theory}, we need to re-evaluate the Kirchhoff stress at each quadrature point in order to apply the operator. This in turn requires evaluating the deformation gradient $\gz F$ and therefore gradients with respect to the undeformed configuration.

\begin{algorithm}[!ht]
  \SetKwInOut{Input}{Given}
  \SetKwInOut{Output}{Return}
  \Input{RHS FE vector $\gz x$,
  cached
  $c_1 := 2\left[\mu - 2 \lambda \log(J)\right]/J$,
  $c_2 := 2\lambda/J$ and $\gz \tau/J$ for each cell and quadrature point,
  evaluated based on $\overline{\gz u^h}$}
  \Output{action of the FE tangent \eqref{eq:algebraic_tangent} on $\gz x$}
  \ForEach{ element $K \in \Omega^h$ }{
        \ForEach{quadrature point $q$ on $K$}{
          evaluate $\gz g := \gz \nabla \gz x$ \tcp*{2nd order}
          evaluate $\boldsymbol{\mathcal{G}}\gz g = \gz g \cdot \left[\gz \tau/J \right]$ \tcp*{2nd order}
          queue $\boldsymbol{\mathcal{G}} \gz g$ for contraction $\gz \nabla \gz N_i : \boldsymbol{\mathcal{G}} \gz g$ \;
          evaluate $\gz g_s := \gz \nabla_{sym} \gz x$ \tcp*{2nd order symmetric}
          evaluate $\boldsymbol{\mathcal{C}} \gz g_s = c_1 \gz g_s + c_2 \, \rm{tr}(\gz g_s) \gz I$ \tcp*{2nd order symmetric}
          queue $\boldsymbol{\mathcal{C}} \gz g_s$ contraction $\gz \nabla_{sym} \gz N_i : \boldsymbol{\mathcal{C}} \gz g_s $ \;
        }
        evaluate queued contractions using sum factorization and distribute results to the destination vector
  }
  \caption{Matrix-free tangent cell operator, variant 1: cache second-order Kirchhoff stress $\gz \tau$ and thereby avoid the need to evaluate refential quantities like $\gz F$ at runtime.}
  \label{alg:mf_tensor2}
\end{algorithm}

In order to avoid recalculation of the Kirchhoff stress, Algorithm \ref{alg:mf_tensor2} caches its value for each element and quadrature point. In this case, we also avoid the need to evaluate gradients of displacement field with respect to the undeformed configuration.
This algorithm also utilizes the chosen constitutive relationship in that the operation of $J \boldsymbol{\mathcal{C}}$ remains directly expressed using \eqref{eq:simplified_action}.
Therefore, the action of the material fourth-order tangent $J \boldsymbol{\mathcal{C}}$ on the second-order symmetric tensor $\gz \nabla_{sym} \gz x$ is cheap to evaluate.
To that end, we cache two scalars that depend on the Jacobian $J$.

\begin{algorithm}[!ht]
  \SetKwInOut{Input}{Given}
  \SetKwInOut{Output}{Return}
  \Input{RHS FE vector $\gz x$,
  cached $\gz \tau/J$ and $\boldsymbol{\mathcal{C}}$ for each cell and quadrature point,
  evaluated based on $\overline{\gz u^h}$}
  \Output{action of the FE tangent \eqref{eq:algebraic_tangent} on $\gz x$}
  \ForEach{ element $K \in \Omega^h$ }{
        \ForEach{quadrature point $q$ on $K$}{
          evaluate $\gz g := \gz \nabla \gz x$ \tcp*{2nd order}
          evaluate $\boldsymbol{\mathcal{G}} \gz g = \gz g \cdot \left[\gz \tau/J\right]$ \tcp*{2nd order}
          queue $\boldsymbol{\mathcal{G}} \gz g$ for contraction $\gz \nabla \gz N_i : \boldsymbol{\mathcal{G}} \gz g$\;
          evaluate $\gz g_s := \gz \nabla_{sym} \gz x$ \tcp*{2nd order symmetric}
          evaluate $\boldsymbol{\mathcal{C}}\gz g_s = \boldsymbol{\mathcal{C}}\gz : \gz g_s$ \tcp*{2nd order symmetric}
          queue $\boldsymbol{\mathcal{C}}\gz g_s$ for contraction $\gz \nabla_{sym} \gz N_i : \boldsymbol{\mathcal{C}}\gz g_s$ \;
        }
        evaluate queued contractions using sum factorization and distribute results to the destination vector
  }
  \caption{Matrix-free tangent cell operator: cache fourth-order material tangent $\boldsymbol{\mathcal{C}}$ and Kirchhoff stress $\gz \tau$.}
  \label{alg:mf_tensor4}
\end{algorithm}

Finally, Algorithm \ref{alg:mf_tensor4} represents the most general case which does not assume any form of the material tangent tensor.
In this case, we have to cache the fourth-order symmetric tensor $\boldsymbol{\mathcal{C}}$ and the Kirchhoff stress $\gz \tau$ for each element and quadrature point, and perform the double contraction with the second-order symmetric tensor $\gz \nabla_{sym} \gz x$ on-the-fly.

@@

For this work we adopt the matrix-free level operators within the GMG and employ Chebyshev pre- and post-smoothing on each level. In general, given a heterogeneous non-linear material, the definition of transfer operators (for restriction and prolongation) and level operators is not trivial. One approach is to start from an arbitrary heterogeneous material at the fine scale and employ homogenization theory \cite{Suquet1985, Hill1972,Hashin1983,Castaneda1997} to design suitable transfer operators \cite{Miehe2007}\footnote{
  We note that this approach is limited to the case where for each quadrature point within an element the fourth-order tangent operator is the same.
}.
In this approach, the fine-scale displacement is decomposed into long-wave and short-wave contributions by an additive split, where the former is associated to the homogeneous response of a patch of elements and the latter represents fluctuations due to the heterogeneous structure of the patch. However, for a patch of elements consisting of the same material, the transfer operators reduce to the standard geometric transfer. Based on this observation we adopt the following approach: It is assumed that the mesh at the coarsest level can accurately describe the heterogeneous finite-strain elastic material. In this case, the restriction and prolongation operations for patches of elements are always performed for the same material and thus admits usage of standard geometric transfer operators. We define the level operators as follows: Recall that the fine scale tangent \eqref{eq:algebraic_tangent} is obtained by linearization around the current displacement field $\gz u^h$. We then restrict this displacement field to all multigrid levels $\{l\}$ and evaluate tangent operators using the matrix-free algorithms from Section \ref{sec:mf}. Essentially, level operators correspond to the linearization around smoothed representation of displacement field\footnote{
Note that this differs from the classical approach where the (tangent) operator $\gz A^{l+1}$ on level $l+1$ is directly related to the (tangent) operator $\gz A^{l}$ on level $l$ via $\gz A^{l+1}=\gz I^{l+1}_{l} \gz A^l \gz I^l_{l+1}$, where $\gz I^l_{l+1}$ and $\gz I^{l+1}_l$ are global prolongation (coarse-to-fine) and restriction (fine-to-coarse) operators, respectively.
}
Finally, at the coarsest level we use a CG solver to reduce the residual by three orders of magnitude within (at most) one hundred steps.
The multigrid algorithm is a V-cycle with two pre- and post-smoothing steps. For other approaches that combine homogenization and multigrid methods for small strain elastic materials, see \cite{Bayreuther2003,Fish1995}.

@@

In this section, we apply the proposed matrix-free operator algorithms for finite-strain hyperelastic material as well as the geometric multigrid preconditioner to a benchmark problem from \cite{Miehe2007}. The coarse mesh for the 2D and 3D problems is illustrated in Figure \ref{fig:miehe}. The 2D material consists of a square (depicted in blue), a hole and two spherical inclusions (depicted in red). The size of the square domain is $10^{-3} \rm{mm}$.
The matrix material is taken to have Poisson's ratio $0.3$ and shear modulus $\mu=0.4225 \times 10^6\, \rm{N/mm^2}$. The inclusion is taken to be $100$ times stiffer.
The 3D material is obtained by extrusion of the 2D geometry into the third dimension.
Note that, although a relatively coarse mesh is used at the coarsest multigrid level, the geometry of the inclusions is captured accurately by manifold descriptions of the boundaries and interfaces.
Both domains are fully fixed along the bottom surface and a distributed load is applied at the top in the $(1,0)$ or $(1,1,0)$ direction for the 2D and the 3D problems, respectively.
The distributed load of $0.5\,\rm{N}$ is applied in 5 steps.
With respect to the nonlinear solver, the displacement tolerance of the $\mathcal{\ell}_2$ norm for the Newton update is taken to be $10^{-5}$ whereas the relative and absolute tolerances for the residual forces is $10^{-8}$. The relative convergence criteria for the linear solver is $10^{-6}$.

\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{0.8\textwidth}
    \centering
    \includegraphics[width=\textwidth]{gmg_2d.png}
    % \caption{2D multigrid mesh with 2 global refinements}
  \end{subfigure}
  \caption{2D multigrid mesh after two global refinements distributed into three MPI processes. Color indicates the process owning the element.}%
  \label{fig:miehe_gmg}
\end{figure}
%
Figure \ref{fig:miehe_gmg} illustrates the hierarchy of multigrid meshes for the 2D problem after two global refinements.

\begin{table}
  \centering
  \begin{tabular}{ccccc}
  \hline
    $p$ & $q$ & $N_{gref}$ & $N_{el}$ & $N_{DoF}$ \\
  \hline
    1 & 2 & 7 & 1441792 & 2887680 \\
    2 & 3 & 6 & 360448 & 2887680 \\
    3 & 4 & 5 & 90112 & 1625088 \\
    4 & 5 & 5 & 90112 & 2887680 \\
    5 & 6 & 5 & 90112 & 4510720 \\
    6 & 7 & 4 & 22528 & 1625088 \\
    7 & 8 & 4 & 22528 & 2211328 \\
    8 & 9 & 4 & 22528 & 2887680 \\
  \hline
  \end{tabular}
  \caption{Parameters for the 2D benchmark: $p$ is the polynomial degree,
  $q$ is the quadrature order, $N_{gref}$ is the number of global mesh refinements, $N_{el}$ is the number of elements and $N_{DoF}$ is the number of DoFs.
  }
  \label{tab:input_parameters_2d}
\end{table}

\begin{table}
  \centering
  \begin{tabular}{ccccc}
  \hline
    $p$ & $q$ & $N_{gref}$ & $N_{el}$ & $N_{DoF}$ \\
  \hline
    1 & 2 & 4 & 1441792 & 4442880 \\
    2 & 3 & 3 & 180224 & 4442880 \\
    3 & 4 & 2 & 22528 & 1891008 \\
    4 & 5 & 2 & 22528 & 4442880 \\
  \hline
  \end{tabular}
  \caption{Parameters for the 3D benchmark: $p$ is the polynomial degree,
  $q$ is the quadrature order, $N_{gref}$ is the number of global mesh refinements, $N_{el}$ is the number of elements and $N_{DoF}$ is the number of DoFs.
  }
  \label{tab:input_parameters_3d}
\end{table}

@@

We are interested in the following metrics:
\begin{itemize}
\item the wall-clock time and memory requirement per DoF for matrix-vector multiplication;
\item the average number of CG iterations throughout the entire simulation (i.e. each Newton-Raphson iteration and each loading step);
\item the wall-clock time per DoF for the CG solver;
\end{itemize}
%
The results of the matrix-free approach with a geometric-multigrid preconditioner are compared to the matrix-based approach in conjunction with the algebraic multigrid (AMG) preconditioner using the package ML of the Trilinos \cite{Heroux2005} library, version 12.12.1.
The aggregation threshold for the AMG preconditioner is taken as $10^{-4}$.
Computations are done on a single node of two clusters:
A node on the ``Emmy'' cluster at RRZE, FAU has two Xeon 2660v2 ``Ivy Bridge'' chips (10 cores per chip + SMT) running at 2.2 GHz with 25 MB shared cache per chip and 64 GB of RAM. Intel's compiler version 18.03 with flags ``-O3 -march=native'' and Intel MPI version 18.03 are used.
The ``IWR'' results were obtained on a single machine with eight Xeon E7-8870 ``Sandy Bridge'' chips (10 cores per chip  + SMT) running at 2.4 GHz with 30 MB shared cache per chip. For the simulation, GCC version 8.1.0 with flags ``-O3 -march=native'' and OpenMPI version 3.0.0 were used.

@@

\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{0.49\textwidth}
      \centering
      \includegraphics[width=\textwidth]{Emmy_RRZE_timing2d.eps}
      \caption{matrix-vector product (2D)}
      \label{fig:benchmark_miehe_Emmy_vmult2}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{Emmy_RRZE_timing3d.eps}
    \caption{matrix-vector product (3D)}
    \label{fig:benchmark_miehe_Emmy_vmult3}
  \end{subfigure}
  ~
  \begin{subfigure}[b]{0.49\textwidth}
      \centering
      \includegraphics[width=\textwidth]{Emmy_RRZE_memory2d.eps}
      \caption{memory consumption (2D)}
      \label{fig:benchmark_miehe_Emmy_memory2}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{Emmy_RRZE_memory3d.eps}
    \caption{memory consumption (3D)}
    \label{fig:benchmark_miehe_Emmy_memory3}
  \end{subfigure}
  \caption{Emmy cluster, RRZE, matrix-vector multiplication.}%
  \label{fig:benchmark_miehe_Emmy}
\end{figure}
